---
title: "Getting started!"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started!}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(noradstats)
```

## Nice to know

-   Write comments: start comment with "\#"
-   Insert sections: use shortcut Ctrl+Shift+R
-   Run code in selected line(s): use shortcut Ctrl+Enter
-   Restart R: use shortcut Ctrl+Shift+F10
-   Save script file: use shortcut Ctrl+S
-   Use forward pipe \|\> : Ctrl+Shift+M

## Load packages

An R package is a collection of functions, data sets, and documentation that extends the capabilities of the core R programming language for specific tasks or analysis.

In this example, we load three packages:

-   noradstats: a Norad package to load and manipulate Norwegian ODA data

-   tidyverse: a collection of R packages that share a common philosophy and syntax for data manipulation and visualization, designed to make data analysis and visualization in R more consistent and user-friendly.

-   janitor: a package to clean and summarise datasets

```{r, eval = FALSE}
# Load packages
library(noradstats)
library(tidyverse)
library(janitor)
```

## Import data

To read statsys data, the .csv data file must be located in the project folder.

```{r, eval=FALSE}
# Read statsys data into a data frame
df_statsys_ten <- read_aiddata("statsys_ten.csv")

# Clean the column names
df_statsys_ten <- clean_names(df_statsys_ten)

# Have a look at the data structure
glimpse(df_statsys_ten)

```

## ODA data

```{r, eval=FALSE}
# Create a data frame including only ODA data and excluding frame agreements
df_oda_ten <- df_statsys_ten |> 
  filter(type_of_flow == "ODA") |> 
  filter(type_of_agreement != "Rammeavtale")

```

## Data manipulation example

```{r, eval=FALSE}
# Create a data frame of ODA by target area in 2021
df_targetarea <- df_oda_ten |> 
  filter(year == max(year)) |> 
  group_by(target_area) |> 
  summarise(nok_mill = sum(disbursed_mill_nok)) |> 
  adorn_totals("row")
```
