---
title: "Access Norwegian and international ODA data"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Access Norwegian and international ODA data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(noradstats)
```

## Tips and tricks

-   Write comments: start comment with "\#"
-   Insert sections: use shortcut Ctrl+Shift+R
-   Run code in selected line(s): use shortcut Ctrl+Enter
-   Restart R: use shortcut Ctrl+Shift+F10
-   Save script file: use shortcut Ctrl+S
-   Use forward pipe \|\> : Ctrl+Shift+M

## First, load packages

An R package is a collection of functions that extends the capabilities of the core R programming language for specific tasks or analysis. In this example, we load three packages:

-   `noradstats`: a Norad package to load and manipulate Norwegian ODA data

-   `tidyverse`: a collection of R packages that share a common philosophy and syntax for data manipulation and visualization, designed to make data analysis and visualization in R more consistent and user-friendly.

-   `janitor`: a package to clean and summarise datasets

```{r, eval = FALSE}
# Load packages
library(noradstats)
library(tidyverse)
library(janitor)
```

## Norwegian development aid statistics
Import current official development assistance data from statsys, inluding columns not publised on bistandsresultater.no

### Import data

Use the `read_aiddata()` function to import statsys data to R. The default path is "Norad-Avd-Kunnskap - Statistikk og analyse/06. Statistikkdatabaser/4. CSV/statsys_total.csv". You can specify the path argument in the function.

```{r, eval=FALSE}
# Read statsys data into a data frame
df_statsys_total <- read_aiddata()

```

### Add basic columns: Norwegian target areas, regions and income categories

The function add_cols_basic adds the following columns:

-   target_area_no: Target area groups in Norwegian, same as on bistandsresultater.no

-   main region_no: Regions in Norwegian, same as on bistandsresultater.no

-   income_category_no: Income categories in Norwegian, same as on bistandsresultater.no

```{r, eval = FALSE}
# Add basic columns
df_statsys_total <- add_cols_basic(df_statsys_total)

```

### Clean column names

Clean column names using the `clean_names` funtion from the `janitor` package to make it easier to refer to the column names. Use the `glimpse` function to get an overview of the dataset.

```{r, eval = FALSE}
# Clean the column names
df_statsys_total <- clean_names(df_statsys_total)

# Have a look at the data structure
glimpse(df_statsys_total)
```

### ODA data only

Filter statsys data to include only ODA and removing frame agreements. The `filter` function is a core function in the `dplyr` package in metapackage `tidyverse`.

```{r, eval=FALSE}
# Create a data frame including only ODA data and excluding frame agreements
df_oda_total <- df_statsys_total |> 
  filter(type_of_flow == "ODA") |> 
  filter(type_of_agreement != "Rammeavtale")

```

### Data manipulation example
A summarised table of the ODA by target area in 2021. Using the functions `group_by` and `summarise` in combination to create grouped summaries.
```{r, eval=FALSE}
# Create a data frame of ODA by target area in 2021
df_targetarea <- df_oda_total |> 
  filter(year == max(year)) |> 
  group_by(target_area) |> 
  summarise(nok_mill = sum(disbursed_mill_nok)) |> 
  adorn_totals("row") |> 
  ungroup()
```

## Norwegian imputed multilateral ODA to countries 
Use the `noradstats` function get_imputed_countries to import Norwegian imputed multilateral ODA to countries. You can specify the startyear and endyear arguments in the function. Run `?get_imputed_countries()` to read the function documentation.

A NOK column is included in the dataset using the OECD exchange rate.

```{r, eval=FALSE}
df_imputed_countries <- noradstats::get_imputed_countries()
glimpse(df_imputed_countries)

```

### Data manipulation example
A table of the top fifty recipient countries of Norwegian imputed multilateral ODA in 2020. Exluding non-countries. Use the `slice_max` function sort and keep only the selected number of observations.
```{r, eval=FALSE}
df_imputed_countries |> 
  filter(obs_time == 2020) |> 
  filter(!str_detect(recipient_label_en, "Total|regional|WB-LM|WB-L|WB-UM|Developing countries, unspecified|Part I Unallocated by income")) |> 
  select(recipient_label_en, nok_mill) |> 
  slice_max(nok_mill, n = 10)
```


## International developmentaid statistics (OECD)
Use the `noradstats` function get_donors to import OECD DAC donor data of ODA to countries and regions. You can specify the startyear and endyear arguments in the function. Run `?get_donors()` to read the function documentation.

```{r, eval = FALSE}
df_donors <- get_donors(startyear = 2020, endyear = 2020)
glimpse(df_donors)
```

### Data manipulation example
A table of the top ten DAC donors of ODA to Uganda in 2020.
```{r, eval=FALSE}

df_donors |> 
  filter(recipient_label_en == "Uganda") |> 
  select(donor_label_en, usd_mill) |> 
  slice_max(usd_mill, n = 10)
```
